---
title: Get started with *corrsurv*
output: 
  github_document:
  html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## About

Collection of two-sample tests for treatment effects with paired censored
survival data and recurrent events survival data. The methods are implementations of 
three papers by Susan Murray and Nabihah Tayob.

## Install

First, install and load the `corrsurv` package from Github using the following 
code:

If you have not installed devtools, run the following:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=F}
install.packages("devtools")
```

If devtools is sucessfully installed, run: 

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
if(!require(corrsurv)) {
  library(devtools)
  install_github('umich-biostatistics/corrsurv') 
}
```

Load the package:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
library(corrsurv)
```

## Summary of methods

This package implements methods from the following three papers by Tayob and Murray:

### Method 1: Murray, 2000

##### Description

Perform two-sample tests for treatment effects with paired censored survival data.

##### Reference

Murray, Susan. Nonparametric Rank-Based Methods for Group Sequential Monitoring of 
Paired Censored Survival Data. 2000. Biometrics, 56, pp. 984-990.

Jump to the `pairtest()` section for the methods from this paper.

### Method 2: Tayob and Murray, 2014

##### Description

Perform the Tayob and Murray two-sample recurrent events test.

##### Reference

Tayob, N. and Murray, S., 2014. Nonparametric tests of treatment 
effect based on combined endpoints for mortality and recurrent events. Biostatistics, 
16(1), pp.73-83.

Jump to the `TM()` section for the methods from this paper. 

### Method 3: Tayob and Murray, 2016

##### Description

Estimate the tau-restricted mean survival across multiple follow-up intervals.

##### Reference

Tayob, N. and Murray, S., 2016. Nonparametric restricted mean analysis across multiple 
follow-up intervals. Statistics & probability letters, 109, pp.152-158.

Jump to the `TM2()` section for the methods from this paper.

## Results

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=F}
vignette("learn-corrsurv")
```
