---
 title: "*Vignette* for `pairsurv`"
 date: "December 20, 2019"
 geometry: margin=2cm
 output: 
  pdf_document: 
    toc: no
 header-includes:
  - \usepackage{booktabs}
  - \usepackage{hyperref}
  - \usepackage{float}
  - \usepackage{caption}
  - \floatplacement{figure}{H}
---

```{r, echo=F, warning=F, message=F, cache=T, include=F, eval=T}
options(digits = 3)
```


## Introduction
 
This vignette presents a step-by-step guide to the analysis of paired survival data
using methods from the following papers:

#### Method 1

Murray, Susan. Nonparametric Rank-Based Methods for Group Sequential Monitoring of 
Paired Censored Survival Data. 2000. Biometrics, 56, pp. 984-990.

Use the `pairtest()` function to use the methods from this paper.

#### Method 2

Tayob, N. and Murray, S., 2014. Nonparametric tests of treatment 
effect based on combined endpoints for mortality and recurrent events. Biostatistics, 
16(1), pp.73-83.

Use the `TM()` function to use the methods from this paper. 

#### Method 3

Tayob, N. and Murray, S., 2016. Nonparametric restricted mean analysis across multiple 
follow-up intervals. Statistics & probability letters, 109, pp.152-158.

Use the `TM2()` function to use the methods from this paper.


## Install
 
First, install and load the `pairsurv` package and other necessary packages:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
if(!require(pairsurv)) {
  library(devtools)
  install_github('umich-biostatistics/pairsurv') 
}
```

## Data

Two simulated data sets are included in the package. The data set `current` is used whenever `curr_method = "current"` and `historical` is used whenever `curr_method = "historical"`. Here, we load the data sets into memory.

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
data("pairdata")
data("pairdata2")
```

## Results

```{r, echo=F, warning=F, message=F, cache=T, include=F, eval=T}
options(digits = 3)
```


#### pairtest(), Murray (2000)

##### Example 1: Analysis of package data set `pairdata`.

Data: 

Enrolled 3711 patients with mild-to-severe nonproliferative or early proliferative 
diabetic retinopathy in both eyes from April 1980 to July 1985 – One eye per patient 
was randomized to early photocoagulation – The other eye deferred photocoagulation 
until detection of high-risk proliferative retinopathy. 
• Survival endpoint: Time to severe visual loss (deﬁned as visual acuity < 5/200 
  at two consecutive visits) subject to censoring 
• eventually detected a beneﬁt with early photocoagulation 
• after 9 years of follow-up, only 5.5% of patients had experienced the event (lots of censoring)

Important Feature of ETDRS Data: Paired Event Times

Paired Design Attractive: 
• Historically used to minimize variability in treatment comparisons 
• With positive correlation in pairs that is taken into account in an analysis: – higher power for ﬁxed sample size than independent designs – smaller required sample size for ﬁxed power than independent designs 
• At the time the ETDRS study was published, the investigators did not have access to these paired tests for censored survival data, but they recognized the statistical issues relating to the correlated structure of the data. Some exploratory analysis on their part suggested “that not taking pairing into account led to conservative tests (ETDRS Research Group, 1991a, p. 749).”


```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
head(pairdata)
```

The function `pairsurv` requires the time-to-event data for each pair (`x1` and `x2`)
and the event indicator for each of the times (`delta1` for `x1`, and `delta2` for `x2`).
The event indicators (`delta1`, `delta2`) equal 1 if the event occured and 0 if the 
patient was censored.

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
eyeresults = pairtest(x1 = pairdata$x1, delta1 = pairdata$delta1, 
                      x2 = pairdata$x2, delta2 = pairdata$delta2, n = 3711)
```

Print results:
```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
print(eyeresults)
```


If you need to access any result not printed by pairtest, use the $ operator to 
search available objects to extract. The test statistic and p-value for each 
test is returned from the call to pairsurv in a list.

Here are the printed and stored results extracted from pairsurv:


```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
# Upper limit of integration
(integration_upper_limit = eyeresults$upperlim)

# Logrank statistic
(logrank_statistic = eyeresults$lgrk.stat.p)
(logrank_statistic_p = 2*(pnorm(logrank_statistic, lower.tail = T))) # since negative statistic

# Gehan statistic
(gehan_statistic = eyeresults$gehan.stat.p)
(gehan_statistic_p = 2*(pnorm(gehan_statistic, lower.tail = T))) # since negative statistic

# Years of Life statistic
(yls_statistic = eyeresults$yls.stat.p)
(yls_statistic_p = 2*(pnorm(yls_statistic, lower.tail = F))) # since positive statistic

# Pepe Fleming statistic
(pepe_flem_statistic = eyeresults$pf.stat.p)
(pepe_flem_statistic_p = 2*(pnorm(pepe_flem_statistic, lower.tail = F))) # since positive statistic

# Logrank statistic assuming independence
(logrank_assuming_indep = eyeresults$lgrk.nopair.stat.p)
(logrank_assuming_indep_p = 2*(pnorm(logrank_assuming_indep, lower.tail = T)) ) # since negative statistic

# Gehan statistic assuming independence
(gehan_assuming_indep = eyeresults$gehan.nopair.stat.p)
(gehan_assuming_indep_p = 2*(pnorm(gehan_assuming_indep, lower.tail = T))) # since negative statistic

# Years of Life assuming independence
(yls_assuming_indep = eyeresults$yls.nopair.stat.p)
(yls_assuming_indep_p = 2*(pnorm(yls_assuming_indep, lower.tail = F))) # since positive statistic

# Pepe Fleming assuming independence
(pf_assuming_indep = eyeresults$pf.nopair.stat.p)
(pf_assuming_indep_p = 2*(pnorm(pf_assuming_indep, lower.tail = F))) # since positive statistic

# group size adjustment
nstuff = ((3711*3711) / (3711 + 3711))^(.5)

# estimate area between survival curves and 95% conf. int.
(yls.diff = eyeresults$yls.num / nstuff)

(yls.upper = yls.diff + 1.96*sqrt(eyeresults$yls.type.var) / nstuff)
(yls.lower = yls.diff - 1.96*sqrt(eyeresults$yls.type.var) / nstuff)

# 95% CI assuming independence
(yls.upper2 = yls.diff + 1.96*sqrt(eyeresults$yls.nopair.type.var) / nstuff)
(yls.lower2 = yls.diff - 1.96*sqrt(eyeresults$yls.nopair.type.var) / nstuff)
```


##### Other options in print and plot methods



#### TM(), Tayob and Murray (2014)

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
head(TMdata)
```

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
N=nrow(TMdata) #Number of subjects
X=TMdata$X
delta=TMdata$delta
table(delta)
#Z=array(cbind(TMdata$Z1,TMdata$Z2,TMdata$Z3,TMdata$Z4,TMdata$Z5,TMdata$Z6),c(N,6))
Z = TMdata[,c('Z1', 'Z2', 'Z3', 'Z4', 'Z5', 'Z6')]
A=max(X) #length of study
Treatment=as.numeric(TMdata$Group==1) #1 if Case and 0 if Control
table(Treatment,delta)
```

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
srec.average = TM(X = X, delta = delta, Z = Z, Group = Treatment, Tau = A/2, 
                  t = seq(from = 0, to = A-A/2, by = A/4))
```

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
print(srec.average)
```

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
srec.area = TM(X = X, delta = delta, Z = Z, Group = Treatment, Tau = A/2, 
               t = seq(from = 0, to = A-A/2, by = A/4), method = "area")
```

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
print(srec.area)
```

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
plot(srec.area)
```


##### Other options in print and plot methods




#### TM2(), Tayob and Murray (2016)

The function `TM2()` allows the user to compute restricted mean survival across multiple follow-up intervals.

The reference paper is Tayob, N. and Murray, S., 2016. Nonparametric restricted mean analysis
 across multiple follow-up intervals. Statistics & probability letters, 109, pp.152-158.
 
This method allows for flexibility in choosing variance estimate for the restricted mean survival. The options
are `proposed`, `sandwich`, and `independence`. See the reference paper for details.

Data: summarize, describe from the data help

The data set contains follow-up times over multiple intervals (X) and the censoring
indicator (delta; 1 if subject died, 0 if censored). The data set is in the "long" format
where multiple events are each recorded as a new row rather than a new column.

The data set contains a total of 100 follow-up times.

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
data("TM2data")
?TM2data
max(TM2data$X)
```

The maximum recorded time in the data set is:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
max(TM2data$X)
```

Similar to the other methods in this package, `TM2` is the work-horse estimation function
and by calling `print()` on the resultant fit object a printed summary of the output will display at the console,
and by calling `plot()` a plot of estimate and 95% bounds on the estimate will be displayed.

An example of analysis with `tau = 12` and all three variance estimates outputed:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
output = TM2(X = TM2data$X, delta = TM2data$delta, Tau = 12, 
             t = seq(from = 0, to = 24, by = 6), var_output = "all")
print(output)
plot(output)
```


An example of analysis with `tau = 12` and proposed variance estimate outputed:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
output = TM2(X = TM2data$X, delta = TM2data$delta, Tau = 12, 
             t = seq(from = 0, to = 24, by = 6))
print(output)
plot(output)
```

An example of analysis with `tau = 12` and independent variance estimate outputed:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
output = TM2(X = TM2data$X, delta = TM2data$delta, Tau = 12, 
             t = seq(from = 0, to = 24, by = 6), var_output = "independence")
print(output)
plot(output)
```

An example of analysis with `tau = 12` and sandwich variance estimate outputed:

```{r, echo=T, warning=F, message=F, cache=T, include=T, eval=T}
output = TM2(X = TM2data$X, delta = TM2data$delta, Tau = 12, 
             t = seq(from = 0, to = 24, by = 6), var_output = "sandwich")

print(output)
plot(output)
```

##### Other options in print and plot methods


